---
title: "Settler colonial projects explain variation in US state formation in the early 20th c."
author: "Frank Edwards"
date: "last updated `r Sys.Date()`"
output: bookdown::html_document2
---
```{r setup, include = F, message = F, cache = F}
knitr::opts_chunk$set(message = F, echo = F, warning = F)
library(tidyverse)
library(lme4)
library(maps)
library(usmap)
library(grid)
library(gridExtra)
library(brms)
library(tidycensus)
library(tigris)
library(units)
library(sf)
library(tidybayes)
library(rstan)

options(tigris_use_cache = T)
options(scipen = 999)
options(dplyr.print_max = 30)
map<-map_data("state")
theme_set(theme_minimal())
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

```{r read, include = F, message = F}
### state shape file
st<-states(cb = T) %>%
  st_transform(8528) %>% 
  filter(!STUSPS%in%
           c("HI", "AK", 
             "PR", "GU", "AS","VI",
             "MP"))

# read data from make_* scripts
pop<-read_csv("./data/temp_population_ts.csv")%>%
  filter(STUSPS!="AK")
disp<-read_csv("./data/temp_disposession_ts.csv")%>%
  filter(STUSPS!="AK")
dist<-read_csv("./data/temp_distribution_ts.csv")%>%
  filter(STUSPS!="AK")
bs<-read_csv("./data/temp_bs_ts.csv")
agents<-read_csv("./data/temp_agents.csv")

## add region to all
disp<-disp %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct()) %>% 
  left_join(bs) %>% 
  left_join(agents)

dist<-dist %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct()) %>% 
  left_join(bs) %>% 
  left_join(agents)
```

```{r run-models, cache = T, include = F, message = F}
#source("models.R")
```

# Introduction

Settler colonialism deeply and unevenly impacts the development of social policy systems. Settler colonial regimes engage in diverse forms of dispossession, extraction, exploitation, and social reproduction. We show that spatial and temporal variation in the salience of settler projects is closely related to the kinds of infrastructure that settler colonial states develop for social control and social reproduction. We show that the formation of the social state in the early 20th century in the US varied dramatically across places as a function of geographic variation in the salience of a set of interests of the settler colonial project: land dispossession, labor exploitation, and reproduction of the settler polity. 

We examine three deeply intertwined, but spatially and temporally variable, contexts for local policymaking in the Progressive Era (1900 - 1930): the Jim Crow agricultural economy; mass urbanization and industrialization; and the rapid distribution of Native lands to white settlers. Each of these contexts presented different sets of problems to local and national settler policy elites. 

Under sharecropping regimes in the US South, white settler elites sought to create and reproduce brutal and paternalistic regimes of informal and formal social regulation that kept Black workers and their families dependent on and subordinate to white interests. Industrial and social white elites in the North sought to discipline large numbers of European immigrants toward nascent ideals of a white American settler polity oriented toward a narrow gendered division of labor and subordination to the interests of industrial capital. In 1862, the federal government enacted the Morill, Homestead, and Pacific Railroad Acts, each of which provided mechanisms for the rapid distribution of dispossessed Native lands to settlers. White settlers and settler corporate interests flooded recently disposessed Native lands with the promise of free or cheap title, seeking to rapidly develop the settler nation in recently admitted states.

We suggest that each of these contexts yielded different trajectories for the formation of US state infrastructure. In the analysis that follows, we leverage data from a range of archival sources to map contexts for early 20th century settler colonization across the US, then build statistical models to evaluate whether these contexts were related to the kinds of capacities that local governments developed for contemporaneous social control and social reproduction. 

Consistent with the historigraphy on the Progressive Era (GORDON, FOX, OTHERS), we find clear relationships between place and policy infrastructure. State formation trajectories varied dramatically across the US, and this variation is closely related to the local salience of distinct racialized settler colonial projects. In places where land speculation and homesteading fueled rapid white settlement, governments and churches hired large numbers of teachers for the intertwined goals of building a white settler polity on the 'frontier' and to continue the coerced assimilation of Indigenous children through a notorious and brutal system of boarding shcools. In places where tenant farming by Black families was common, there was relatively little investment in state bureaucracies. Social control and reproduction instead relied on a coordinated formal and informal system of racial terror that developed alongside a relatively weak set of state social policy institutions.

The foundations upon which states built out their infrastructure to control, educate, and care for their populations were closely related to regional variation in the political economy of settler colonialism across regions in the US. States developed more or less robust institutions in various domains to address the problems of governance posed by the local projects of settler colonization that both elites and ordinary settlers sought to advance. 

We focus our attention on three interconnected settler colonial structures and processes that had large impacts on social life in the early 20th century US: 1) the active dispossession and erasure of Native nations and the development of settler enterprises and communities; 2) the exploitation of Black labor for agricultural production and the fortification of an apartheid social order; and 3) the coercive and paternalistic assimilation of recent immigrants from Eastern and Southern Europe into an urban industrial settler colonial polity. We argue that these deeply related, but geographically uneven, processes were related to the development of state capacities to resolve emerging problems of domestic governance in a period of rapid social transformation. Or, put more directly, that THE LOCAL SALIENCE OF PARTICULAR CONSITUENTS / STRUCTURES IS RELATED TO STATE FORMATION. 

These are three of the projects that were critical to settler colonialism in the early 20th c US. Not the only ones of course! Colonial processes deeply impacted the experiences of a broad range of groups and produced a range of distinct racialized social structures (CITES FOR MEXICO BORDER, MIGRANT WORKERS, CHINESE EXCLUSION, JAPANESE INTERNMENT, PHILLIPINES COLONIZATION, AND OTHERS). *Here, we narrowly focus on three settler colonial projects to illustrate how a macro-comparative approach can be helpful in validating core findings in historical scholarship and in incorporating the insights of critical scholarship on settler colonization and racial capitalism into some of the core methods of political, economic, and comparative historical sociology REVISE.*

# Three settler colonial projects in the early 20th c. US

The geographic distribution of the local implementation of the three focal settler projects we focus on is crudely illustrated in Figure \@ref(fig:three-maps). The federal government was most heavily involved in actively redistributing Native lands to settlers in states in the Midwest, Great Plains, Mountain West, and on the West Coast. While Native land disposession was still deeply salient in Eastern states, questions of settler land ownership and the establishment of coherent settler polities had been resolved (albeit incompletely) prior to the late 19th and early 20th centuries. In panel 1, we show the number of acres that had been dispossessed from Native nations and distributed to settlers and settler institutions through two key laws that heavily impacted patterns of colonial settlement in the late 19th and early 20th centuries, the Morill and Homestead acts of 1862. 

The Jim Crow system of apartheid was developed and becoming deeply entrenched across the states in the US South that had built elaborate settler plantation economies through the exploitation and terrorization of formerly enslaved Black people. As of 1910, the first wave of the Great Migration of African Americans out of the US South was underway, but, as shown in Panel 2 of Figure \@ref(fig:three-maps), the Black population in the US was still heavily concentrated in the US Southeast. 

And the Industrial Revolution, coupled with a wave of crises across Eastern and Southern Europe, created both strong push and pull forces for large populations of immigrants to travel to North America. These immigrants settled across the industrial cities of the Northeast, Midwest, and West coasts. Many of them also joined the waves of white settlers pursuing the opportunity to claim recently dispossessed Native lands in the Central and Western portions of the US. As shown in Panel 3 of Figure \@ref(fig:three-maps), the populations of many states in the Northeast, Midwest, and West Coast were becomingly increasingly urban and foreign born. 

Below, we briefly summarize these three social, economic, and demographic contexts for US state formation in the early 20th century, a period routinely referred to as the 'Progressive Era'. During this period, rapid increases in the complexity, capacity, and scope of domestic governance were occurring at the federal, state, and local levels. Following this brief discussion of historical contexts, we turn to a discussion and analysis of how these varying contexts for settler colonization related to heterogeneity in the kinds of governments that settlers built across places.

```{r three-maps, fig.cap = "Geographic and demographic contexts in 1910 for settler colonial governance"}
### ADD DISPOSESSION
p1 <- dist %>% 
  select(STUSPS, year,
         morill_acres_cum, homestead_acres_cum,
         total_acres_2021) %>% 
  filter(year == 1910) %>% 
  mutate(dist = morill_acres_cum + homestead_acres_cum) %>% 
  left_join(st) %>% 
  ggplot(aes(fill = dist / 1e6,
             geometry = geometry)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Acres\n(millions)",
       subtitle = "1. Lands patented through\nMorill and Homestead acts")

p2 <- st %>% 
  left_join(pop %>% 
              filter(year == 1910)) %>% 
  ggplot(aes(fill = blk / tot_pop)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Proportion",
       subtitle = "2. Black population")

p3 <- st %>% 
  left_join(pop %>% 
              filter(year == 1910)) %>% 
  ggplot(aes(fill = urban_immig_pop / tot_pop)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Proportion",
       subtitle = "3. Urban immigrant population")

grid.arrange(p1, p2, p3, nrow = 1)
```

## Settling the frontier

OVERVIEW: BUILDING THE SETTLER NATION STATE, ERASING NATIVE NATIONS

Dispossession

Redistribution

THE MORILL, HOMESTEAD AND RR ACTS

Nation building

Forced assimilation

```{r fig.cap = "Land cessions from Native nations claimed by US government 1795 - 1893"}
st<-states(cb = T) %>%
  st_transform(8528) %>% 
  filter(!STUSPS%in%
           c("HI", "AK", 
             "PR", "GU", "AS","VI",
             "MP"))

st_acres<-data.frame(state = st$STUSPS,
                     total_acres_2021 = as.numeric(set_units(st_area(st), "acre")))

### royce map data
## available here: https://data-usfs.hub.arcgis.com/datasets/usfs::tribal-lands-ceded-to-the-united-states-feature-layer/about
## and archived here: https://data.nativeland.info/dataset/indian-land-cessions-in-the-united-states-1784-1894/resource/e2b874d5-3c39-4287-bf89-d4ef14b87c83
cessions<-read_sf("./data/us_forest_cessions_royce/S_USA.TRIBALCEDEDLANDS.shp")

cessions_dat<-read_csv("./data/us_forest_cessions_royce/royce.csv") %>% 
  mutate(time = as_date(TribalCededLandsTableCessDate1),
         year = year(time)) %>% 
  rename(CESSNUM = TribalCededLandsNewCESSNUM) %>% 
  select(CESSNUM, year)

cessions<-cessions %>% 
  left_join(cessions_dat)

cessions<-cessions %>% 
  st_transform(8528)

cessions_diff<-st_difference(st, st_union(cessions))

cessions_diff<-cessions_diff %>% 
  mutate(year = NA)

### layer state boundaries
cessions<-st_intersection(cessions, st)
### add in settler claimed territories as of 1796
cessions2 <- cessions %>%  
  bind_rows(cessions_diff %>%  
              mutate(year = NA))

ggplot(cessions2,
       aes(fill = year)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  labs(fill = "Year") 
```

MANY OF THE LOCATIONS THESE FIRST GENERATION SETTLERS WERE GOING TO WERE RECENT DISPOSESSIONS, PARTICULARLY CIVIL WAR ERA. IN 1900, MANY OF THESE LANDS HAD BEEN CLAIMED BY US IN VERY RECENT PAST, WITHIN 50 YEARS. SECOND / THIRD WAVE OF WESTERN SETTLERS. BUILDING THE SETTLER COLONIAL POLITY / NATION STATE AS KEY AGENDAS.


```{r fig.cap = "Disposession of Native lands in the 19th c."}
p1<-ggplot(disp,
       aes(x = year,
           group = STUSPS)) + 
  geom_line(aes(y = cumulative_ceded / 1e6)) + 
  labs(x = "Year",
       y = "Acres (millions)") + 
  lims(y = c(0, 150)) + 
  facet_wrap(region~division) + 
  labs(subtitle = "1. Total area")

p2<-ggplot(disp,
       aes(x = year,
           group = STUSPS)) + 
  geom_line(aes(y = prop_ceded)) + 
  labs(x = "Year",
       y = "Proportion") + 
  facet_wrap(region~division) + 
  labs(subtitle = "2. Proportion of state area")

grid.arrange(p1, p2, ncol = 2)
```

This shows the timing and spatial distribution in cumulative land disposession by state and region. Panel 1 shows the lands claimed by the US in millions of acres by state from 1796 - 1894. Panel 2 shows these patterns of Native land loss as a proportion of the contemporary areas of US states.  

Clear regional patterns here. Land loss occurred earlier in states closer to the Atlantic coast. Dispossession accelerated in the North Central and Western regions throughout the middle portion of the 19th c. 

```{r fig.cap = "Lands patented under two legislative mechanisms for the distribution of disposessed Native Lands to settlers, cuulative total 1930"}
pdat<-disp %>% 
  select(STUSPS, total_acres_2021) %>% 
  distinct() %>% 
  left_join(dist %>% 
              filter(year == 1930) %>% 
              select(STUSPS, morill_acres_cum, homestead_acres_cum)) %>% 
  replace_na(list(morill_acres_cum = 0, homestead_acres_cum = 0)) %>% 
  left_join(st)

p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = morill_acres_cum / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Acres (millions)",
       subtitle = "1. Morill Act") + 
  theme(legend.position = "bottom")


p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = homestead_acres_cum / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Acres (millions)",
       subtitle = "2. Homestead Act") + 
  theme(legend.position = "bottom")

grid.arrange(p1, p2, ncol = 2)
```


## Entrenching the exploitation of Black labor and marginalization of Black social and political power

Between 1900 and 1930, settler elites in the the US South were busy entrenching Jim Crow and profiting from the deeply exploitative sharecropping system of agricultural production.

Figure \@ref(fig:jim-crow-maps) shows two key demographic features of this period. In panel 1, we illustrate the entrenchment and ubiquity of the sharecropping system of racially exploitative agrictultural production. In panel 2, we show that, unlike much of the rest of the United States at the time, the white population of the US South was composed overwhelmingly of white settlers who had been born in the United States. 

### The sharecropping economy

DEFINE SYSTEM

SUMMARIZE LITERATURE

### The white southern polity

SUMMARIZE LIT ON SOUTHERN WHITENESS AND COHERENCE OF WHITE SETTLER POLITY IN US SOUTH

### The early and mid-career of Jim Crow

Racial order, law, and terrorism

ESP THINK ON BUREAUCRACY, INFORMALITY OF SOCIAL CONTROL, COLLABORATION BETWEEN FORMAL AND INFORMAL, VIGILANTISM AND LAW

EDUCATION AND JIM CROW

SOCIAL WORK AND JIM CROW

```{r jim-crow-maps, fig.cap = "Black tenant farming and immigrant status of white settlers, 1910"}
pdat<-st %>% 
  left_join(pop) %>% 
  filter(year == 1910)

p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = blk_sharecrop / 1e5)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(hundred thousands)",
       subtitle = "1. Black population living on tenant farms") + 
  theme(legend.position = "bottom")

p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = white_immig / (white_immig + white_non_immig))) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Proportion",
        subtitle = "2. Proportion of white poulation foreign born") + 
  theme(legend.position = "bottom")

grid.arrange(p1, p2, nrow = 1)

```


## Incorporating urban white immigrants into the settler polity and industrial working class

The industrialization of the US Northeast and Midwest during the late 19th and early 20th centuries spurred two related demographic processes: rapid urbanization and mass European immigration. Figure \@ref(fig:immig-maps) displays the distribution of both the urban population and the white immigrant population in the US in 1910. States like New York, Massachusetts, Pennsylvania, Ohio, and Illinois saw dramatic growth in their populations, with much of that growth composed of European immigrants and concentrated in cities like Chicago, Boston, Cleveland, Philadelphia, and New York. 

```{r immig-maps, fig.cap = "White immigrant population and urban population, 1910"}
pdat<-st %>% 
  left_join(pop) %>% 
  filter(year == 1910)

p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = white_immig / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(millions)",
       subtitle = "1. White, foreign-born") + 
  theme(legend.position = "bottom")

p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = urban_pop / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(millions)",
        subtitle = "2. Urban") + 
  theme(legend.position = "bottom")

grid.arrange(p1, p2, nrow = 1)
```

SUMMARIZE DEMOGRAPHY LIT

### The rise of progressive domestic governance in Northern US cities

social work

policing

## Patterns in the development of state infrastructure 1900 - 1930

state capacity is hard to measure, but lets try! Here we are going to use employment measures for three roles that came to play critical roles in the developing practices of social welfare and social control: teachers, police, and social workers. 

We also think about physical infrastructure for similar sets of institutions. We look at various forms of institutional group quarters: correctional institutions, mental institutions, and institutions for the poor and disabled.

Indian Boarding schools are another key piece of physical infrastructure. We look at those too, but I am still waiting on validating some of that data. 

### Employment in education, policing, and social welfare: 1900 - 1930

There was dramatic growth in public sector employment in general (CHECK THIS), and in roles targeted at nation building, social control, and social welfare in particular during the Progressive Era. *These dramatic increases in employment rates were reflected in the increasing complexity and reach of a wide range of public institutions described by a number of historians (CITES).*

In Figure \@ref(fig:emp-ts), we display employment rates for the lower 48 US States between 1900 and 1930. Employment of teachers increased from a rate of 5.5 per 1,000 population in 1900 to a rate of 8.6 per 1,000 in 1930, a 56 percent increase. Employment of police doubled between during this period, increasing from a rate of about 0.7 per 1,000 in 1900 to a rate of about 1.4 per 1,000 in 1930. Outside of a few cities in the northeast, social workers were almost never recorded in the Census in 1900 (only 361 recorded nationally in 1900). However, by 1930, the Census recorded about 17,000 social workers nationally, and increase of over 40 fold in terms of employment rates per capita. 

```{r emp-ts, fig.cap = "Employment in three core social policy professions over time, US rates per 1,000 population, 1900 - 1930"}
emp_pdat<-pop %>% 
  group_by(year) %>% 
  summarize(pop = sum(tot_pop),
            Teachers = sum(teachers),
            Police = sum(police),
            social_workers = sum(socwork)) %>% 
  pivot_longer(cols = Teachers:social_workers) %>% 
  mutate(name = case_when(
    name == "Teachers"~ "1. Teachers",
    name == "Police" ~ "2. Police",
    name == "social_workers" ~ "3. Social Workers")) %>% 
  mutate(rate = value / pop * 1e3)

emp_pdat  %>% 
  ggplot(aes(x = year, y = rate)) + 
  geom_line() + 
  facet_wrap(~name, scales = "free_y") + 
  labs(x = "Year", y = "Employment rate")

```

However, this expansion of the reach of the state did not occur uniformly across the country. Figure \@ref(fig:emp-maps) displays employment rates per 1,000 population in 1930 for teachers, police, and social workers for each of the lower 48 US States. Employment rates for teachers were lowest-in-the nation in the US South. Employment rates for teachers were highest in a cluster of Great Plains states, like the Dakotas, Wyoming, Nebraska, Iowa, and Kansas. Employment rates for police were at their peak in 1930 in Northeastern, Midwestern, and West Coast states like New York, New Jersey, Massachusetts, Illinois, and California. Police employment rates were relatively lower in places like Arkansas, Missisippi, and North Carolina. The distribution of social workers in 1930 shares some features with the distribution of police: states in the Northeast, Midwest, and West coast like New York, Ohio, Massachusetts, California, and Minnesota had high rates of social worker employment, while states in the Great Plains and South, like Missisippi and South Dakota recorded very few social workers in the 1930 Census. 

```{r emp-maps, fig.cap = "Employment in three core social policy professions across US states in 1930, rates per 1,000 population"}
pdat<-st %>% 
  left_join(pop) %>% 
  filter(year == 1930)

p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = teachers / tot_pop * 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Rate",
       subtitle = "1. Teachers") + 
  theme(legend.position = "bottom")

p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = police / tot_pop * 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Rate",
        subtitle = "2. Police") + 
  theme(legend.position = "bottom")

p3<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = socwork / tot_pop * 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Rate",
        subtitle = "3. Social Workers") + 
  theme(legend.position = "bottom")

grid.arrange(p1, p2, p3, nrow = 1)
```

### Institutional capacity: prisons, jails, asylums, orphanages and poorhouses

State capacity could also be measured through the ability of states to engage in therapeutic, correctional, or paternalistic confinement. In Figure \@ref(fig:gq-map) we show the distribution of the population housed in institutional group quarters in 1930 across US states. Using the 1930 Census, we display the population housed in prisons in jails in panel 1, the population housed in mental institutions in panel 2, and the population housed in institutions for the poor and/or disabled in figure 3. Using data from the Department of the Interior's ``Federal Indian Boarding School Initiative'' (2022), we display a count of Federal Indian boarding schools operating in 1930 in panel 4. I MAY ADD ENROLLMENT DATA FROM ARCIA REPORTS ONCE WE VALIDATE IT. 

Data on institutional group quarters populations are less reliable than employment data for constructing a full 1900 - 1930 Census time series. 

shows Prisons in NE, S, CA. Mental and poverty instutitons mostly in Urban NE, MW, CA

```{r gq-map, fig.cap = "Institutional group quarters, US 1930"}
pdat<-st %>% 
  left_join(pop) %>% 
  left_join(bs) %>% 
  filter(year == 1930)

p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = inst_correctional / 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(thousands)",
       subtitle = "1. Correctional institutions") + 
  theme(legend.position = "bottom")

p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = inst_mental / 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(thousands)",
        subtitle = "2. Mental institutions") + 
  theme(legend.position = "bottom")

p3<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = inst_poor_disabled / 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(thousands)",
        subtitle = "3. Institutions for the\npoor and disabled") + 
  theme(legend.position = "bottom")

grid.arrange(p1, p2, p3, nrow = 2)
```

### Institutional capacity: Indian boarding schools

FROM DOI REPORTS

Figure \@ref{fig:bsts} displays the number of Indian boarding schools operating annually between 1830 and 1930, with eras of Federal Indian law shaded for reference. The number of boarding schools grew dramatically during the Reservation and Assimilation eras, with most growth occurring between 1870 and 1905. After 1905, a slow decline in the number of operating schools began continued until the end of our focal period, 1930. At the beginning of the Removal Era (1830), there were 18 boarding schools in operation, and 23 were operating at the end of the Removal era (1850). During the subsequent Reservation era (1850 - 1887), the number of operating boarding schools increased from 23 to 173. The number of operating facilities continued to increase during the Assimilation era to a maximum of 242 in 1908, followed by a decline to 174 facilities recorded as operating in 1930, a similar level of facilities operating as had existed at the beginning of the period. 

```{r bsts, fig.cap = "Indian boarding schools operating annually, US 1830 - 1930"}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

t<-bs %>% 
         filter(year<=1930,
                year>=1830) %>% 
         group_by(year) %>% 
         summarize(boarding_schools = sum(boarding_schools))

ggplot(t) +
       aes(y = boarding_schools,
           x = year) + 
  geom_line() + 
  annotate("rect",  # add removal era shading
           ymin = 0, 
           ymax = 250,
           xmin = 1830,
           xmax = 1850,
           color = NA,
           fill = cbPalette[1],
           alpha = 0.2) + 
  annotate("rect",  # add reservation era shading
           ymin = 0, 
           ymax = 250,
           xmin = 1850,
           xmax = 1887,
           color = NA,
           fill = cbPalette[2],
           alpha = 0.2) + 
    annotate("rect",  # add reservation era shading
           ymin = 0, 
           ymax = 250,
           xmin = 1887,
           xmax = 1930,
           color = NA,
           fill = cbPalette[3],
           alpha = 0.2) + 
  annotate("text",
           x = c(1840, 1860, 1910),
           y = c(60, 90, 160),
           color = cbPalette[1:3],
           label = c("Removal era", "Reservation era", "Assimilation era")) +
  labs(x = "Year", y = "Number of facilities") + 
  scale_x_continuous(breaks = seq(1830, 1930, by = 20),
                     limits = c(1830, 1930)) 
  
```

In Figure \@ref{fig:bsts2} we zoom in on our focal period. This uses two sources of data, p1 is DoI, p2 is ARCIA. EXPLAIN THAT THESE ARE LIKELY UNDERCOUNTS AGAIN.

Expansion of program via enrollment even as N facilities is declining and capacity is declining. surprise surprise. 

```{r bsts2, fig.cap = "Boarding schools and recorded student enrollment and capacity, US totals 1890 - 1920"}
## map out the enrollment and capacity data
bs_ts<-agents %>% 
  left_join(bs) %>% 
  left_join(st)

t<-bs_ts %>% 
  group_by(year) %>% 
  summarize(Capacity = sum(boarding_n_acc),
            Enrollment = sum(students_enroll)) %>% 
  pivot_longer(cols = Capacity:Enrollment) 

### for in-text callouts
p1<-ggplot(t, 
           aes(x = year, y = value, color = name)) + 
  geom_line() + 
  labs(x = "Year", y = "Children",
       color = "",
       subtitle = "2. Reported capacity and enrollment (ARCIA)")

p2<-ggplot(bs_ts %>% 
  group_by(year) %>% 
  summarize(boarding_schools = sum(boarding_schools)),
           aes(x = year, y = boarding_schools)) + 
  geom_line() + 
  labs(x = "Year", y = "Count",
       subtitle = "1. Facilities operating annually (Dept. of Interior)")

grid.arrange(p2, p1, nrow = 1) 
```

Figure \@ref{fig:bsmap} shows the geographic distribution of boarding school infrastructure and utilization when enrollments were at a 1890 - 1920 period maximum in 1917. DISCUSS

```{r bsmap, fig.cap = "Boarding schools in operation and reported enrollment and capacity at peak enrollment (1917)"}
p3<-ggplot(bs_ts %>% 
         filter(year == 1917),
       aes(fill = students_enroll / 1e3,
           geometry = geometry)) + 
  geom_sf() + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Children\n(thousands)",
       subtitle = "3. Reported enrollment") + 
  theme(legend.position = "bottom")

p2<-ggplot(bs_ts %>% 
         filter(year == 1917),
       aes(fill = boarding_n_acc / 1e3,
           geometry = geometry)) + 
  geom_sf() + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Children\n(thousands)",
       subtitle = "3. Reported capacity") + 
  theme(legend.position = "bottom")

p1<-ggplot(bs_ts %>% 
         filter(year == 1917),
       aes(fill = boarding_schools,
           geometry = geometry)) + 
  geom_sf() + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Facilities",
       subtitle = "1. Boarding schools operating") + 
  theme(legend.position = "bottom")

grid.arrange(p1, p2, p3, nrow = 1)
```

# Analytic strategy

The works of settler colonial and racial politics historical and theoretical scholars suggest a straightforward approach to selecting measures that are either a) direct, or b) proxy measures for elite white settler interests given a few assumptions about relationships between population distributions and political processes.

To explore the relationship between settler colonial political and economic contexts and state formation, we evaluate whether a set of policy outcomes are related to a set of demographic and economic predictors at the state and year-levels. We conduct exploratory and inferential analysis using data on US states between 1900 and 1930. 

DESCRIBE THE MEASURES IN GENERAL TERMS HERE

Generally our approach suggests that state formation in a particular place *s* and time *t* is a function of settler interests specific to that time and place. 

\[E[\textrm{infrastructure}_{st}] = f(\textrm{settler interests}_{st})\]

We proceed by producing and evaluating descriptive data visuals for our operational measures of 'settler interests' and comparing distributions of these measures across states and over time. We then produce and evaluate a similar set of visuals to evaluate the distribution of our 'infrastructure' measures. Finally, we estimate regression models that evaluate the relationships between our measures of infrastructure and settler interests, conditional on time-stable place variation, national temporal trends, and basic characteristics of the population in the state and year.

## Data

All measures on features of the US population are derived from the 1900, 1910, 1920, and 1930 decennial US Census 100% samples, accessed through IPUMS (Ruggles et al.). For each census year and for each of the lower 48 US states, we count the total population, the child population (as the population under 18 years old), the white urban immigrant population (as the population identified as foreign born, white, and non-Hispanic living in an urban household), and the Black population living in tenant farm households. 

State-year employment measures are computed from the 1900 - 1930 Census. A census respondent is counted as a police employee if their code for the IPUMS variable OCC1950 is equal to one of {771 'Marshals and constables', 773 'Policemen and detectives', 782 'Sheriffs and bailiffs'}; is counted as a social worker if OCC1950 is coded 079 'Social and welfare worker'; and counted as a teacher if OCC1950 is equal to 093 'Teachers (n.e.c)', which excludes music and art teachers but includes other primary and secondary teachers. 

We also compute measures of populations in institutional group quarters, but rely on a slightly different set of samples. The 1910 census did not collect information on group quarters type. We rely on the 1900, 1920, and 1930 100% samples for the group quarters population by type, and the 1910 1.4% sample which contains IPUMS imputations for group quarters type. From these samples, we derive state-year measures of the population housed in correctional instiuttions, the population housed in mental institutions, and the population housed in institutions for the elderly, handicapped, and poor. 

Indigenous land dispossession over time is measured using shape files created by the US Forest Service from a set of 1899 maps detailing the boundaries of land cessions claimed by the US government from Native nations through treaties or acts of Congress (CITE). The distribution of Native lands to settlers is measured through data on the implementation of the Homestead Act and Morill Act of 1862. We use data on parcels of land transferred to land-grant universities under the Morill Act as documented by Robert Lee an High Country News as part of their 'Land Grab University' project. These data are compiled from land patents archived by the Bureau of Land Management and represent the most extensive archive of such land transfers currently available. Data on parcels of land distributed to settlers through the Homestead Act are obtained from  the report ``Homesteads,'' by the Bureau of Land Management of the U.S. Department of the Interior, 1962, and digitized and distributed by Edwards, Friefeld and and Wingo 2017. For both Homestead and Morill act data, we measure cumulative acres parceled since 1862 under each act within the state by year. 

## A note on settler colonial geography

geography itself endogenous to settler colonization, this is tricky! we are trying to understand the emergence of the state as bureaucratic entity in this period, so it makes sense to do this, but it is messy! we don't want to imply a teleological reading here. borders are unstable and contested.

## statistical methods and models

Nearly of our 'infrastructure' outcomes are measured as counts of persons. As such, we rely on Negative Binomial likelihoods that can flexibly handle over-dispersed count outcomes to model these employment and population processes. For facility counts (boarding schools), we employ a similar model that removes population offsets from the specification. 

DISCUSSIONS OF MODEL SELECTION IN APPX

Model 1. is

\[ y_{st} \sim \textrm{NegBin}(\lambda, \alpha)\]

\[ \lambda = \textrm{E}[y_{st}] \]

\[ \log \left( \frac{\textrm{E}[y_{st}]}{\textrm{population}_{st}} \right) = \gamma_s + \beta_{0t} + \beta_1 \frac{\textrm{Black tenant farmers}_{st}}{\textrm{population}_{st}} + \beta_2 \frac{\textrm{White urban immigrants}_{st}}{\textrm{population}_{st}} + \beta_3 \textrm{Morill and Homestead acres}_{st} + \beta_4 \log(\textrm{child population}_{st}) \]

\[\gamma_s \sim N(0, \sigma^2_\gamma)\]

With weakly informative priors

\[\beta \sim N(0, 20)\]
\[\sigma \sim t(3, 0, 2.5)\]
\[\alpha \sim \textrm{Gamma}(0.01, 0.01)\]

Model 2 (for zero inflated outcomes) is

XXXXX

# Findings

We suggest that regionally-specific political contexts, driven by spatially and temporally varying interests of white settler elites, caused variation in trajectories of welfare state formation across US states. In this study, we explore how three settler colonial political projects mattered for the emergence of the US welfare state: the economic exploitation and social exclusion of Black people from the settler polity, the consolidation of settler political, social, and cultural claims on Indigenous territories, and the incorporation of urban European immigrants into the settler polity. 

## Employment models


```{r}
# # ### extract posterior draws for b1, b2, b3, set up for plotting
# # pdat<-teachers_m1_bayes %>% 
# #   gather_draws(c(b_scaleIblk_sharecropDtot_pop,       
# #                  b_scaleIurban_immig_popDtot_pop,             
# #                  b_scalemorill_acres_cumPhomestead_acres_cum)) %>% 
# #   mutate(model = "1. Teachers") %>% 
# #   bind_rows(
# #     cops_m1_bayes %>% 
# #       gather_draws(c(b_scaleIblk_sharecropDtot_pop,       
# #                      b_scaleIurban_immig_popDtot_pop,             
# #                      b_scalemorill_acres_cumPhomestead_acres_cum)) %>% 
# #       mutate(model = "2. Police")) %>% 
# #   bind_rows(
# #     socwork_m1_bayes %>% 
# #       gather_draws(c(b_scaleIblk_sharecropDtot_pop,       
# #                      b_scaleIurban_immig_popDtot_pop,             
# #                      b_scalemorill_acres_cumPhomestead_acres_cum)) %>%
# #       mutate(model = "3. Social workers")) %>% 
# #   mutate(.variable = case_when(
# #     .variable == "b_scaleIblk_sharecropDtot_pop" ~ 
# #       "β1: Black tenant\nfarmers per capita",
# #     .variable == "b_scaleIurban_immig_popDtot_pop" ~ 
# #       "β2: Urban white\nimmigrants\nper capita",
# #     .variable == "b_scalemorill_acres_cumPhomestead_acres_cum" ~ 
# #       "β3: Cumulative \nMorill +\nHomestead acres"))
# 
# 
# ggplot(pdat,
#        aes(x = .value, 
#            y = .variable)) +
#   stat_halfeye() + 
#   geom_vline(aes(xintercept = 0), lty = 2) + 
#   labs(x = "Estimate",
#        y = "") + 
#   facet_wrap(~model) + 
#   theme_bw()
```

## Group quarters models



RAN BOARDING SCHOOL MODEL WITH DISCONTINUITY ON ACREAGE, IT WORKS WELL. 
