---
title: "Simultaneous Settler Projects and U.S. Racial State Formation
METHODS AND FINDINGS"
author: "Frank Edwards and Theresa Rocha Beardall"
date: "last updated `r Sys.Date()`"
output: bookdown::pdf_document2
---
```{r setup, include = F, message = F, cache = F}
knitr::opts_chunk$set(message = F, echo = F, warning = F)
library(tidyverse)
library(lme4)
library(maps)
library(usmap)
library(grid)
library(gridExtra)
library(brms)
library(tidycensus)
library(tigris)
library(units)
library(sf)
library(tidybayes)
library(rstan)
library(knitr)
# library(trackdown)
# # trackdown::trackdown_auth_configure(path = "client_secret_243891325977-kgvbep6m6eb8t2feb8rvu9fneo8d6msa.apps.googleusercontent.com.json")
# # trackdown::update_file(file = "ms.Rmd", 
# #                        hide_code = T,
# #                        rich_text = T)
options(tigris_use_cache = T)
options(scipen = 999)
options(dplyr.print_max = 30)
map<-map_data("state")
theme_set(theme_minimal())
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
# set hook for thousands comma
knit_hooks$set(inline = function(x) {
prettyNum(round(x,2), big.mark=",")
})
```

```{r read, include = F, message = F}
### state shape file
st<-states(cb = T) %>%
  st_transform(8528) %>% 
  filter(!STUSPS%in%
           c("HI", "AK", 
             "PR", "GU", "AS","VI",
             "MP"))
# read data from make_* scripts
pop<-read_csv("./data/population_ts_full.csv")%>%
  filter(STUSPS!="AK")
disp<-read_csv("./data/temp_disposession_ts.csv")%>%
  filter(STUSPS!="AK")
dist<-read_csv("./data/temp_distribution_ts.csv")%>%
  filter(STUSPS!="AK")
bs<-read_csv("./data/temp_bs_ts.csv")
## add region to all
disp<-disp %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct()) %>% 
  left_join(bs) 
dist<-dist %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct()) %>% 
  left_join(bs) 
```

```{r run-models, cache = T, include = F, message = F}

mdat<-pop %>% 
  left_join(dist) %>% 
  left_join(bs) %>% 
  mutate(blk_sharecrop_pc = blk_sharecrop / tot_pop,
         urban_immig_pc = urban_immig_pop / tot_pop,
         homestead_plus_morill = homestead_acres_cum + morill_acres_cum)

# rescale year
mdat<-mdat %>% 
  mutate(year_c = (year - 1900)/10)

# source("models.R")

teachers_m1<-readRDS("./models/teachers_m1.rds")
cops_m1<-readRDS("./models/cops_m1.rds")
socwork_m1<-readRDS("./models/socwork_m1.rds")
prison_m1<-readRDS("./models/prison_m1.rds")
mental_m1<-readRDS("./models/inst_m1.rds")
poor_m1<-readRDS("./models/poor_m1.rds")
bs_m1_z<-readRDS("./models/bs_m1_z.rds")
```

# Analytic strategy

We describe the relationship between settler colonial political and economic contexts and state formation by evaluating whether a set of measures capturing social policy infrastructure are related to a set of measures that are related to locally-specific settler colonial policy contexts at the state and year-levels. We use data on US states between 1900 and 1930 derived from US Census and archival sources. 

While there have been a number of compelling analyses of the long-run effects of variation in colonial systems on colonized societies (ACEMOGLU ETC) and of chattel slavery and racial terror on contemporary politics (ACHARYA) we refrain from engaging in formal causal inference in this analysis. Rich theoretical and historical work suggests that racial and settler colonial processes deeply structure policy outcomes through complex and long-running processes that become obscured and entangled with a battery of social processes. We suggest that these processes are visible at a relatively high-level in the historical data. Settler colonialism is endogenous to the whole of US politics, economics, and demography. Our interest here is not to identify causal effects, but rather to descriptively chart how the historical geography of settler colonialism in the US overlaps with the historical geography of US state formation. Typical approaches in causal inference seek to eliminate stable cross-sectional differences in units. Our approach prioritizes understanding both these stable cross-sectional differences and dynamics of change over time. Toward this end, we pursue a descriptive and interpretive macro-sociological approach to understanding how settler contexts related to heterogeneity in state formation in the early 20th century. 

Our approach is theory-driven, informed by insights derived from 1) RACIAL STATE and SETTLER COLONIAL THEORIES OF US STATE FORMATION and 2) the HISTORIOGRAPHY OF EACH REGION IN THIS PERIOD. We suggest that RACIAL STATE THEORY provides a coherent framework for understanding heterogeneity in state formation during this period, by applying the simultaneous settler projects framework to these distinct contexts. 

Generally our approach suggests that social policy infrastructure in a particular place *s* and time *t* is a function of settler elite interests specific to that time and place. 

$E[\textrm{infrastructure}_{st}] = f(\textrm{settler interests}_{st})$

We prioritize a descriptive approach that illustrates differences in macro-social contexts across places and their associations with state-formation outcomes and trajectories. We describe and evaluate the geographic and temporal co-occurrence of state formation outcomes and discrete settler projects. 

We proceed by producing and evaluating descriptive data visuals and maps of our operational measures of settler colonial policy contexts and comparing distributions of these measures across states and over time. We then produce and evaluate a similar set of visuals to evaluate the distribution of our social policy infrastructure measures. We pay particular attention to regional patterns in these distributions, and situate our interpretations in dialogue with prior historical and theoretical research. 

Finally, we estimate regression models that evaluate the relationships between our measures of infrastructure and settler interests, conditional on time-stable place variation, national and regional temporal trends, and basic characteristics of the population in the state and year. These models formally evaluate associations between the geographies of settler colonial projects and a range of policy outcomes. 

A replication package is available at github.com/f-edwards/racial_politics_path

## Data

This study leverages historical data from a range of sources. We rely on data from the decennial US Census (1900 - 1930) for information on population features, employment, and group quarters populations. Information on the timing and location of Indigenous land disposession is obtained from shape files constructed from maps of Indigenous land dispossession through treaties and acts of congress compiled by the US Congress in 1899 (Royce Maps). Data on the distribution of Native lands to white settlers under the authority of the Homestead Act (1862) are obtained from federal archives of land patents. Data on the distribution of Native lands to US State governments for the construction of land-grant universities under the authority of the Morill Act (1862) was compiled by High Country News from federal archives (CITE). Finally, we leverage data on Indian Boarding Schools compiled by The Department of the Interior under their Federal Indian Boarding School Initiative (CITE 2022).

### Population composition and employment 

All measures on key features of the US population are derived from the 1900, 1910, 1920, and 1930 decennial US Census 100% samples, accessed through IPUMS (Ruggles et al.). We use complete decennial census microdata for each decennial year to reduce the potential of measurement error for relatively small state-level sub-populations of interest. For each census year and for each of the lower 48 US states, we count the total population, the child population (as the population under 18 years old), the white non-immigrant population (as the total population identified as non-Hispanic white with both parents 'native-born'), the white immigrant population (as the population identified as non-Hispanic white either foreign born or with at least one foreign born parent), the white urban immigrant population (as the white immigrant population living in a household identified as urban), and the Black population as the total population identified as Black, and the Black tenant farm population as the total population identified as Black living in households identified as farms and where the ownership of the dwelling is identified as rented.

State-year employment measures are also computed from the 1900 - 1930 Census 100 percent microdata files using the variable 'Occupation, 1950 basis' harmonized for earlier Census years by IPUMS staff. A census respondent is counted as a police employee if their occupation is coded as any of the following categories 'Marshals and constables'; 'Policemen and detectives'; or 'Sheriffs and bailiffs'. A respondent is counted as a social worker if their occupation is coded as 'Social and welfare worker'; and counted as a teacher if coded as 'Teachers (n.e.c)', which excludes music, art, and post-secondary teachers but includes other primary and secondary teachers. We compute counts for each of these employment categories for each state and decennial Census year.

### Group quarters populations

We also compute measures of populations in institutional group quarters, but rely on a different set of US Census samples. The 1910 census does not include information on group quarters type. For these data, we rely on the 1900, 1920, and 1930 100% samples for the group quarters population by type, and the 1910 1.4% sample which contains IPUMS imputations for group quarters type. From these samples, we derive state-year measures of the population housed in correctional institutions; the population housed in mental institutions and the population housed in institutions for the elderly, handicapped, and poor (a single category in Census files). Because recorded populations housed in institutions for the elderly, handicapped, and poor are  sparse in some places and years, we combine these counts with populations hosed in mental institutions. This sum of populations housed in paternalistic institutions serves as a proxy measure of the development of what has been described as the 'therapeutic' state (CITE POLSKY) or the 'white parental state' (CITE WARD), or the precursors to the modern institutions of the welfare state (CITE KATZ). 

### Indigenous land disposession and settler redistribution

Indigenous land dispossession over time is measured using shape files created by the US Forest Service from a set of 1899 maps detailing the boundaries of land cessions claimed by the US government from Native nations through treaties or acts of Congress known as the 'Royce Maps', which detail lands claimed by the US Government from Native nations by either treaty or act of Congress  (CITE LOC U.S. Serial Set, Numbers 4014 and 4015). These maps provide information on the timing and location of Indigenous dispossession by the US government between 1784 and 1894, which we use to visualize the historical geography of settler westward expansion in the late 19th century. 

The redistribution of Native lands to settlers by the US federal government is measured through data on the implementation of the Homestead and Morrill Acts of 1862. We use data on parcels of land transferred to land-grant universities under the Morrill Act as documented by Robert Lee and High Country News as part of their 'Land Grab University' project. These data are compiled from land patents archived by the Bureau of Land Management and represent the most extensive archive of such land transfers currently available (CITE LGU REPORT). Data on parcels of land distributed to settlers through the Homestead Act are obtained from the report ``Homesteads,'' by the Bureau of Land Management of the U.S. Department of the Interior, 1962, and digitized and distributed by Edwards, Friefeld and and Wingo (2017). For both Homestead and Morill act data, we measure cumulative acres of Native lands parceled since 1862 under each act within each state by year. 

## Statistical methods and models

We develop a series of regression models to explore the conditional relationships between outcomes measuring the capacity of settler state governments and predictors measuring the contexts for three focal settler colonial projects. These state-year models evaluate how three regionally specific settler projects -- Jim Crow, Native land redistribution, and white immigrant assimilation -- related to patterns in the development of core state infrastructure and capacity during a period of rapid increases in state capacity in the US. 

Our first set of models (see equation XXX) predict infrastructure outcomes for each decennial Census year between 1900 and 1930, and include time-stable state-level random intercepts, a non-linear time component (a national decade-specific intercept), and parameters for per capita rates of Black tenant farming, per capita rates of white urban immigrants, and cumulative acreage redistributed to settlers under the Homestead and Morill acts. We use Bayesian Hamiltonian Monte Carlo methods to estimate all models with weakly informative priors on all model parameters. This approach adjusts for national time trends (secular increases in employment across places), as well as adjusting for idiosyncratic differences across states. Note that the log link function of our negative binomial likelihoods implies multiplicative relationships between additive focal predictors and outcomes, resulting in rates of linear change that vary depending on starting values (intercepts). 

We discuss our process for model selection in detail in the appendix.

Most of our infrastructure outcomes are measured as counts of persons. As such, we rely on Negative Binomial likelihoods that can flexibly handle over-dispersed count outcomes to model these employment and population processes. The population offset term on the left-hand side of equation XXXX allows us to adjust outcomes and estimated relationships for the size of state populations (as per capita rates). 

**(1) Regression model for count outcomes**

$$y \sim \textrm{NegBin}(\lambda, \alpha)$$

$$\lambda = \textrm{E}[y]$$

$$\log \left( \frac{\textrm{E}[y]}{\textrm{population}} \right) = \gamma_s + \beta \textrm{X}_{st}$$

$$x_{st} = t \times \left(\frac{\textrm{Black tenant farmers}_{st}}{\textrm{population}_{st}} +  \frac{\textrm{White urban immigrants}_{st}}{\textrm{population}_{st}} +\textrm{Morill and Homestead acres}_{st}\right)$$

$$\gamma \sim N(0, \sigma^2_\gamma)$$

With weakly informative priors

$$\beta \sim N(0, 20)$$

$$\sigma \sim t(3, 0, 2.5)$$

$$\alpha \sim \textrm{Gamma}(0.01, 0.01)$$

We also model the presence and number of Indian boarding schools in each state as a focal measure of state infrastructure in the Progressive era. Unlike other outcomes in our study, there are a substantial number of states that did not have Indian boarding schools during this period. We treat the presence of boarding schools as a zero-inflated process, and use a two-stage model to evaluate their distribution. One component of the model estimates the probability of having at least one boarding school during the focal period (a binary outcome). This binary measure of boarding schools is constant over the period 1900 - 1930 at the state-level, so we estimate this model using data only for the first observation in the period, 1900, following the model in equation (2). The second component of the model estimates the count of boarding schools present in a state-year, conditional on there being at least one boarding school in the state, following equation (1), excluding the population offset term. 

**(2) Regression model for binary outcomes**

$$ y > 0 \sim \textrm{Bernoulli}\left(\theta\right) $$
$$ \textrm{logit} (\theta) =  \beta_{0t} + \beta_1 \frac{\textrm{Black tenant farmers}_{s}}{\textrm{population}_{s}} + \beta_2 \frac{\textrm{White urban immigrants}_{s}}{\textrm{population}_{s}} + \beta_3 \textrm{Morill and Homestead acres}_{s} $$
$$\beta \sim N(0, 20)$$

For all regression models, we transform predictors to a mean-centered standard deviation scale (*z* scores) to improve model convergence in a context of widely differing scales across measures. More details on model selection are provided in the supplemental materials. 

# Findings

We suggest that regionally-specific political contexts, driven by spatially and temporally varying interests of white settler elites, caused variation in trajectories of welfare state formation across US states. In this study, we explore how three settler colonial political projects mattered for the emergence of the US welfare state: the economic exploitation and social exclusion of Black people from the settler polity, the consolidation of settler political, social, and cultural claims on Indigenous territories, and the incorporation of urban European immigrants into the settler polity. 

Below, we briefly summarize these three social, economic, and demographic contexts for US state formation in the early 20th century, a period routinely referred to as the 'Progressive Era'. During this period, rapid increases in the complexity, capacity, and scope of domestic governance were occurring at the federal, state, and local levels. Following this brief discussion of historical contexts, we turn to a discussion and analysis of how these varying contexts for settler colonization related to heterogeneity in the kinds of governments that settlers built across places.

## The geography of settler colonial projects in the late 19th and early 20th century US

Figures \@ref(fig:cessions-map) shows the temporal and spatial process of the dispossession of lands from Native Nations by the US Government between 1795 and 1893 (DATA FROM USFS VIA Lib Of Cong). These maps show a clear difference in the timing of land cessions as a function of place. By 1795, the US Government claimed nearly all lands East of the Appalachian Mountains, including large parts of contemporary Kentucky and Tennessee. Land dispossession throughout the South and the Great Lakes regions accelerated quickly during the early 19th century. Throughout the middle of the 19th century, there were large-scale efforts by the federal government and vigilante settlers to dispossess tribal lands in California, the Midwest, Great Plains, and Mountain West. By 1895, the overwhelming majority of lands in the contemporary lower 48 US states had been claimed by the US federal government. 

```{r cessions-map, fig.cap = "Land cessions from Native nations claimed by US government 1795 - 1893"}
st<-states(cb = T) %>%
  st_transform(8528) %>% 
  filter(!STUSPS%in%
           c("HI", "AK", 
             "PR", "GU", "AS","VI",
             "MP"))
st_acres<-data.frame(state = st$STUSPS,
                     total_acres_2021 = as.numeric(set_units(st_area(st), "acre")))
### royce map data
## available here: https://data-usfs.hub.arcgis.com/datasets/usfs::tribal-lands-ceded-to-the-united-states-feature-layer/about
## and archived here: https://data.nativeland.info/dataset/indian-land-cessions-in-the-united-states-1784-1894/resource/e2b874d5-3c39-4287-bf89-d4ef14b87c83
cessions<-read_sf("./data/us_forest_cessions_royce/S_USA.TRIBALCEDEDLANDS.shp")
cessions_dat<-read_csv("./data/us_forest_cessions_royce/royce.csv") %>% 
  mutate(time = as_date(TribalCededLandsTableCessDate1),
         year = year(time)) %>% 
  rename(CESSNUM = TribalCededLandsNewCESSNUM) %>% 
  select(CESSNUM, year)
cessions<-cessions %>% 
  left_join(cessions_dat)
cessions<-cessions %>% 
  st_transform(8528)
cessions_diff<-st_difference(st, st_union(cessions))
cessions_diff<-cessions_diff %>% 
  mutate(year = NA)
### layer state boundaries
cessions<-st_intersection(cessions, st)
### add in settler claimed territories as of 1796
cessions2 <- cessions %>%  
  bind_rows(cessions_diff %>%  
              mutate(year = NA))
ggplot(cessions2,
       aes(fill = year)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  labs(fill = "Year") 
```

The 1900 geographic distribution of the three focal settler projects we focus on is illustrated in Figure \@ref(fig:three-maps). This set of maps shows illustrates the clear differences in the settler political economy that were most salient in distinct regions of the early 20th century US. 

The extensive and methodical dispossession of Native lands that characterized the Assimilation and Allotment era was in full force during the Progressive Era. Panel 1 of Figure \@ref(fig:three-maps) shows the cumulative area of dispossessed Native lands that had been redistributed to settlers and US states at low or no cost under the Homestead and Morill Acts of 1862 by 1930. Nearly 90 million acres of Native lands had been granted to settlers and states through the Morill and Homestead acts by 1900. By 1930, this total would reach over 265 million acres. By 1930, 47 percent of the contemporary area of Nebraska, and 44 percent of the contemporary area of North Dakota had been granted to settlers under these programs. More than 30 percent of the contemporary area of South Dakota, Oklahoma, Colorado, and Montana had been granted to settlers under these programs by 1930. These programs primarily impacted lands in the US West, Southwest, Mountain West, and Midwest during this period. 

```{r three-maps, fig.cap = "Three focal contexts for settler colonial governance, 1930"}
### ADD DISPOSESSION
p1 <- dist %>% 
  select(STUSPS, year,
         morill_acres_cum, homestead_acres_cum,
         total_acres_2021) %>% 
  filter(year == 1930) %>% 
  mutate(dist = morill_acres_cum + homestead_acres_cum) %>% 
  left_join(st) %>% 
  ggplot(aes(fill = dist / 1e6,
             geometry = geometry)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Acres\n(millions)",
       subtitle = "1. Lands patented through\nMorill and Homestead acts")
p2 <- st %>% 
  left_join(pop %>% 
              filter(year == 1930)) %>% 
  ggplot(aes(fill = blk_sharecrop / tot_pop * 100)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "% of\npop.",
       subtitle = "2. Percent Black \ntenant farmers")
p3 <- st %>% 
  left_join(pop %>% 
              filter(year == 1930)) %>% 
  ggplot(aes(fill = urban_immig_pop / tot_pop* 100)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "% of\npop.",
       subtitle = "3. Percent white \nurban immigrant")

grid.arrange(p1, p2, p3, nrow = 1)
```

The Jim Crow system of apartheid and its economic complement the sharecropping system were deeply rooted across the states in the US South during the Progressive era. Panel 2 of Figure \@ref(fig:three-maps) shows the entrenchment of the sharecropping system in 1930, a dominant mode of racialized political economy in the US South several decades into the first Great Migration of African Americans out of the US South to industrial cities in the Northeast and Midwest. In 1930, 644,000 Black residents of Mississippi were living on tenant farms, 32 percent of the state's total population. More than 400,000 Black people were recorded as residents of tenant farms in Georgia, Alabama, and South Carolina in 1930. More than 300,000 were enumerated in North Carolina, Louisiana, and Texas.  And more than 100,000 were enumerated in Arkansas, Tennessee, and Virginia. 

And the Industrial Revolution, coupled with a wave of crises across Europe, created both strong push and pull forces for large populations of immigrants to travel to North America. Southern and Eastern European immigrants settled across the industrial cities of the Northeast, Midwest, and West coasts. Many of them also joined the waves of white settlers pursuing the opportunity to claim recently dispossessed Native lands in the Central and Western portions of the US. As shown in Panel 3 of Figure \@ref(fig:three-maps), the populations of many states in the Northeast, Midwest, and West Coast had become increasingly urban and foreign born. Nearly 3 million residents of New York state's 12.5 million residents in 1930 were either first or second generation European immigrants (CONFIRM THIS IS URBAN). In Illinois, 1.1 million of the state's 7.6 million residents were first or second generation immigrants from Europe. In Massachusetts and Rhode Island, 23 percent of the each state's population were European immigrants in 1930. More than 10 percent of the population were European immigrants in New Jersey, Connecticut, Michigan, New Hampshire, California, and Pennsylvania.

## The geography of social policy infrastructure 1900 - 1930

We define 'social state capacity' as the resources available to governments to direct and intervene in processes of social control and social reproduction (see Cingolani 2013 for a review of theory and measurement of the more general concept 'state capacity'). Capacity is a function of both physical infrastructure and human resources available to state actors. During the Progressive Era, federal, state, and local governments developed their abilities to intervene into and manage social life through a series of increasingly professionalized bureaucracies. We suggest that these nascent components of the carceral and welfare states developed in markedly different ways across places. To capture some of this heterogeneity, we focus our analysis on employment and infrastructure outcomes in a broad set of social policy systems including education, social welfare, and criminal justice. 

We rely on three sets of measures associated with the capacity of states to accomplish social policy goals: 1) employment rates in key social policy positions as a measure of the relative size of bureaucracies and magnitude of human effort available to policy actors, 2) populations housed in institutional group quarters as a measure of the scope of particular kinds of physical infrastructure used by the welfare and carceral systems during the Progressive Era, and 3) counts of operating Indian boarding schools as indicators of physical infrastructure directed at coerced assimilation and erasure of Indigenous peoples. Taken together, these three sets of measures provide a diverse and broad set of insights into spending on the basic infrastructure of social governance during the early 20th century. 

### Employment in education, policing, and social welfare: 1900 - 1930

There was dramatic growth in public sector employment in general, and in roles targeted at nation building, social control, and social welfare in particular during the Progressive Era (CITATIONS! CHECK NAIDU). *These dramatic increases in employment rates were reflected in the increasing complexity and reach of a wide range of public institutions described by a number of historians (GENERAL LIT ON GOVT COMPLEXITY AND ARTICULATION 1900 - 1930).*

In Figure \@ref(fig:emp-ts), we display employment rates for the lower 48 US States between 1900 and 1930 in panel 1, and the proportional change in employment rates by category over time in panel 2. Employment of teachers increased from a rate of 5.5 per 1,000 population in 1900 to a rate of 8.6 per 1,000 in 1930, a 56 percent increase. Employment of police doubled during this period, increasing from a rate of about 0.7 per 1,000 in 1900 to a rate of about 1.4 per 1,000 in 1930. Outside of a few cities in the northeast, social workers were almost never recorded in the Census in 1900 (only 361 recorded nationally in 1900). By 1930, the Census recorded about 17,000 social workers nationally, an increase of about 30 fold in terms of employment rates per capita. However, this expansion of the reach of state and state-aligned street-level bureaucrats into social life did not occur uniformly across the country. Figure \@ref(fig:emp-maps) displays employment rates per 1,000 population in 1930 for teachers, police, and social workers for each of the lower 48 US States. 

```{r emp-ts, fig.cap = "Employment in three core social policy professions over time, US rates per 1,000 population, 1900 - 1930"}
emp_pdat<-pop %>% 
  group_by(year) %>% 
  summarize(pop = sum(tot_pop),
            Teachers = sum(teachers),
            Police = sum(police),
            social_workers = sum(socwork)) %>% 
  pivot_longer(cols = Teachers:social_workers) %>% 
  mutate(name = case_when(
    name == "Teachers"~ "Teachers",
    name == "Police" ~ "Police",
    name == "social_workers" ~ "Social Workers")) %>% 
  mutate(rate = value / pop * 1e3)
### convert to proportional change
emp_pdat00<-emp_pdat %>% 
  filter(year == 1900) %>% 
  rename(rate00 = rate) %>% 
  select(-year, -value, -pop)
emp_pdat_prop <- emp_pdat %>% 
  left_join(emp_pdat00) %>% 
  mutate(prop_change = rate / rate00) 
p1<-emp_pdat_prop  %>% 
  ggplot(aes(x = year, y = rate, color = name)) + 
  geom_point() + 
  geom_line() + 
  labs(x = "Year", y = "Rate per 1,000",
       color = "",
       subtitle = "1. Employment rate") + 
  theme(legend.position = "bottom")
p2<-emp_pdat_prop  %>% 
  ggplot(aes(x = year, y = prop_change, color = name)) + 
  geom_point() + 
  geom_line() + 
  labs(x = "Year", y = "Proportional change (log scale)",
       color = "",
       subtitle = "2. Change relative to 1900") + 
  scale_y_log10() + 
  theme(legend.position = "bottom")
grid.arrange(p1, p2, nrow = 1)
```

Employment rates for teachers were highest in a cluster of Midwestern and Mountain West states. Thirteen states had teacher employment rates of over 10 per 1,000 population, all in the Midwest or West. North and South Dakota had highest-in-the-nation teacher employment rates, at over 14 per 1,000 persons. Teachers were instrumental in building human capital and national culture in settler homesteading communities. They also served as front-line %>% of the assimilationist programs throughout the region's vast network of Indian boarding schools (Marquez 2024, OTHERS). Employment rates for teachers were lowest-in-the nation in the US South. There were more than twice as many teachers per capita in North Dakota in 1930 than there were in Alabama. In Alabama, Arkansas, Kentucky, Missisippi, and Georgia, the 1930 Census recorded fewer than 7 teacher per 1,000 residents. Southern states in general systematically underfunded or failed to develop public institutions that could benefit African Americans, and the Jim Crow school system was an exemplar of this practice (CITE). 

Employment rates for police in 1930 were highest in Northeastern, Midwestern, and West Coast states with large urban hubs. New York had the highest numbers of recorded police, at nearly 30,000 identified in the 1930 Census (about 2.4 per 1,000 population). New Jersey and Massachusetts also employed more than 2 officers per 1000 population in 1930. The NYPD, and other big-city police agencies had undertook efforts at professionalization and bureaucratization during the Progressive Era (e.g. Guarigilia, OTHERS), resulting in substantial growth in many of the places with large police organizations. By contrast, the largely rural South under a public/private hybrid model of racial terror and segregation during Jim Crow employed relatively fewer police officers. In Arkansas, Mississippi, North Carolina, and Alabama, there were fewer than 0.6 per 1,000 police recorded in the 1930 Census. 

```{r emp-maps, fig.cap = "Employment in three core social policy professions across US states in 1930, rates per 1,000 population"}
pdat<-st %>% 
  left_join(pop) %>% 
  filter(year == 1930)

p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = teachers / tot_pop * 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Rate per\n 1000 pop.",
       subtitle = "1. Teachers") + 
  theme(legend.position = "bottom")
p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = police / tot_pop * 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "",
        subtitle = "2. Police") + 
  theme(legend.position = "bottom")
p3<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = socwork / tot_pop * 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "",
        subtitle = "3. Social Workers") + 
  theme(legend.position = "bottom")
grid.arrange(p1, p2, p3, nrow = 1)

### text callouts
temp<-pop %>% 
  filter(year == 1930) %>% 
  mutate(police_pc = police / tot_pop * 1e3,
         teachers_pc = teachers / tot_pop * 1e3,
         socwork_pc = socwork / tot_pop * 1e3) %>% 
  select(STUSPS, police:socwork, police_pc:socwork_pc) %>% 
  arrange(desc(teachers_pc))
  
### make a prop change map? or delta rate map?
  
```

The distribution of social workers in 1930 shares some features with the distribution of police: states with large cities in the Northeast, Midwest, and West coast like New York, Ohio, Massachusetts, California, and Minnesota had high rates of social worker employment. These places were among the first to see the emergence of core Progressive social welfare institutions like settlement houses and juvenile courts, and saw rapid growth in social work employment. By 1930, five states were employing at least 0.2 social workers per 1,000 population: Massachusetts, New York, California, Rhode Island, and Ohio. As with education, the South exhbited very low levels of social worker employment during this period. In Mississippi, only 24 social workers were recorded in the 1930 Census, an employment rate of about 1.2 per 100,000 population. Social workers were also rare in many of the states with high levels of homesteading like the Dakotas and Oklahoma. 

### Physical infrastructure: prisons, jails, asylums, orphanages and poorhouses

US welfare and carceral systems have historically prioritized developing physical infrastructure to confine and/or treat populations understood to be socially problematic (CITES ON CONFINEMENT). In Figure \@ref(fig:gq-map) we show the distribution of the population housed in institutional group quarters in 1930 across US states as recorded by the US Census to provide one measure of physical infrastructure for social welfare and social control. We display the population housed in prisons in jails in panel 1, the population housed in mental institutions in panel 2, and the population housed in institutions for the poor and/or disabled in figure 3. Note that alternative measures of group quarters populations are available, but choice of measures does not have a meaningful impact on our substantive conclusions (for a detailed discussion, see Appendix).

```{r gq-map, fig.cap = "Population confined in prisons and jails, mental institutions, and institutions for the poor and disabled, 1930"}
pdat<-st %>% 
  left_join(pop) %>% 
  left_join(bs) %>% 
  filter(year == 1930) %>% 
  mutate(inst_correctional_pc = inst_correctional / tot_pop * 1e3,
         inst_mental_pc = inst_mental / tot_pop * 1e3,
         inst_poor_disabled_pc = inst_poor_disabled / tot_pop * 1e3)

p1<-ggplot(pdat) +
  geom_sf(aes(geometry = geometry,
              fill = inst_correctional_pc)) +
  scale_fill_viridis_c() +
  theme_void() +
  labs(fill = "Rate per\n1,000 pop.",
       subtitle = "1. Correctional") +
  theme(legend.position = "bottom")
p2<-ggplot(pdat) +
  geom_sf(aes(geometry = geometry,
              fill = inst_mental_pc)) +
  scale_fill_viridis_c() +
  theme_void() +
  labs(fill = "",
        subtitle = "2. Mental") +
  theme(legend.position = "bottom")
p3<-ggplot(pdat) +
  geom_sf(aes(geometry = geometry,
              fill = inst_poor_disabled_pc)) +
  scale_fill_viridis_c() +
  theme_void() +
  labs(fill = "",
        subtitle = "3. Poor and disabled") +
  theme(legend.position = "bottom")
grid.arrange(p1, p2, p3, nrow = 1)
# for in-text callouts
pdat<-st %>% 
  left_join(pop) %>% 
  left_join(bs) %>% 
  filter(year == 1930)%>% 
  mutate(inst_correctional_pc = inst_correctional/tot_pop * 1e3,
         inst_mental_pc = inst_mental / tot_pop * 1e3,
         inst_poor_disabled_pc = inst_poor_disabled / tot_pop * 1e3) 
pdat_callout<-pdat %>% 
  pivot_longer(cols = inst_correctional_pc:inst_poor_disabled_pc,
               names_to = "inst_type",
               values_to = "rate") %>% 
  group_by(inst_type) %>% 
  summarize(mean = mean(rate),
            sd = sd(rate)) %>% 
  as_tibble() %>% 
  select(-geometry)

temp<-pdat %>% 
  ungroup() %>% 
  select(-geometry) %>% 
  select(STUSPS, inst_correctional_pc:inst_poor_disabled_pc)
  
```

Panel 3 displays the rate of confinement in institutions for the poor and disabled across US states in 1930. These institutions included poorhouses, settlement homes, orphan asylums, sanitariums, and similar institutions targeted at containing and/or supporting social groups understood to be 'dependent' on state or religious charity support. We find that rates of confinement were lowest in states in the South, like Arkansas, Alabama, Mississippi and Florida. New England, the mid-Atlantic, and Great Lakes states had relatively high rates of confinement in these kinds of institutions. Delaware, Wisconsin, and Massachusetts led the nation in populations housed in institutions for the poor and disabled, each recording at least 5 residents per 1,000 population in 1930. By contrast, these institutions were less developed in the South. Arkansas, Alabama, Florida, Georgia, Texas, and Mississippi each recorded about 1 resident in institutions for the poor and disabled per 1,000 population. 

Panel 2 displays regional patterns in confinement in institutions for the mentally ill across places in 1930. States in the mid Atlantic, New England, and West Coast had the highest rates of confinement, at over 3 per 1,000 on average. In Vermont, Massachusetts, New York, and New Hampshire confinement rates in 1930  exceeded 4 per 1,000 population. States in the East South Central Division (Mississippi, Alabama, Tennessee, and Kentucky) had the lowest average levels of confinement, at 1.4 per 1,000 persons. Levels of confinement ranged between 2 and 3 on average across the rest of the country. Of note, we observe no inmates in mental institutions in North Dakota in the 1930 census data.

Panel 1 of Figure \@ref{fig:gq-map} shows state-level variation in confinement in correctional group quarters reported in the 1930 US Census. We find average confinement rates to be lowest in the Mid-Atlantic and New England states, at 1.2 per 1,000 population. Confinement rates were highest on average in the West Coast states (2.5 per 1,000). Average rates across the remaining regions varied between 1.3 and 2.0 per 1,000 population. Kansas, home to federal and military prisons in Leavenworth, had the highest incarceration rates in the nation, at 4.5 persons per 1,000 population.

Whereas states in the South tended to have very low levels of confinement in institutions for the poor and disabled and relatively low levels of confinement in mental institutions, they had levels of confinement in correctional institutions that were near national averages. This suggests that Southern policy makers and philanthropic allies systematically invested fewer resources in physical infrastructure targeted at containing or supporting the poor, young, mentally ill, and disabled than did other states, while maintaining and developing prisons and jails at rates similar to the rest of the nation. DISCUSS WARD BRIEFLY HERE. 

### Physical infrastructure: Indian boarding schools

Indian boarding schools were built by the US government and its allies in to eliminate Native nations through the coercive assimilation of Indigenous children (Adams 1995; Hoxie 2001). Investment in Indian boarding schools peaked during the Assimilation and Allotment era of Federal Indian Law, roughly spanning 1887 - 1934. Notably, these carceral and genocidal institutions represented one of the earliest sustained investments by the federal government in education infrastructure (alongside the Morill act of 1862). Below, we provide a descriptive overview of the geography and timing of boarding school operations. Following initiatives by Secretary of the Interior Deb Haaland and leaders from across Indian Country, the US Department of the Interior has initiated its first comprehensive review of the scope and impacts of the Indian boarding school program. We rely on data produced by the Department of the Interior (NEWLAND 2022) for historical counts of boarding schools in operation by state and year. 

Figure \@ref(fig:bsts) displays the number of Indian boarding schools operating annually between 1830 and 1930, with eras of Federal Indian law shaded for reference. The number of boarding schools grew dramatically during the Reservation and Assimilation eras, with most growth occurring between 1870 and 1905. After 1905, a slow decline in the number of operating schools began and continued until the end of our focal period, 1930. At the beginning of the Removal Era (1830), there were 18 boarding schools in operation, and 23 were operating at the end of the Removal era (1850). During the subsequent Reservation era (1850 - 1887), the number of operating boarding schools increased from 23 to 173. The number of operating facilities continued to increase during the Assimilation era to a maximum of 242 in 1908, followed by a decline to 174 facilities recorded as operating in 1930, a similar level of facilities operating as at the beginning of the period. 

```{r bsts, fig.cap = "Indian boarding schools operating annually, US 1830 - 1930"}
cbPalette <- c("#999999", "#E69F00", "#56B4E9")
t<-bs %>% 
         filter(year<=1930,
                year>=1830) %>% 
         group_by(year) %>% 
         summarize(boarding_schools = sum(boarding_schools))
ggplot(t) +
       aes(y = boarding_schools,
           x = year) + 
  geom_line() + 
  annotate("rect",  # add removal era shading
           ymin = 0, 
           ymax = 250,
           xmin = 1830,
           xmax = 1850,
           color = NA,
           fill = cbPalette[1],
           alpha = 0.2) + 
  annotate("rect",  # add reservation era shading
           ymin = 0, 
           ymax = 250,
           xmin = 1850,
           xmax = 1887,
           color = NA,
           fill = cbPalette[2],
           alpha = 0.2) + 
    annotate("rect",  # add reservation era shading
           ymin = 0, 
           ymax = 250,
           xmin = 1887,
           xmax = 1930,
           color = NA,
           fill = cbPalette[3],
           alpha = 0.2) + 
  annotate("text",
           x = c(1840, 1860, 1910),
           y = c(60, 90, 160),
           color = cbPalette[1:3],
           label = c("Removal era", "Reservation era", "Assimilation\nand allotment era")) +
  labs(x = "Year", y = "Number of facilities") + 
  scale_x_continuous(breaks = seq(1830, 1930, by = 20),
                     limits = c(1830, 1930)) 
  
```

Figure \@ref(fig:bsmap) shows the geographic distribution of boarding schools in operation when enrollments were at a 1830 - 1930 period maximum (242 facilities documented as operating in 1907). The Department of the Interior has identified Indian Boarding schools operating in 1907 in most regions of the country, with the exception of New England and the Southeast. There were at least 64 boarding school facilities operating in Oklahama in 1907, by far the highest number in the US. There were 8 states with at least 10 facilities in operation in 1907: South Dakota (24), Minnesota (18), Arizona (14), Montana (14), New Mexico (11), Washington (11), Alaska (10, not displayed), and Wisconsin (10). There were at least 5 boarding schools in operation in 7 states: California (7), North Dakota (7), Oregon (7), Colorado (5), Hawaii (5, not displayed), Idaho (5), and Kansas (5). States with at least one confirmed facility in operation in 1907 included Nebraska (3), Nevada (3), Pennsylvania (3), Utah (3), Wyoming (3), Iowa (2), Michigan (2), Arkansas (1), Illinois (1), Missouri (1), New York (1), North Carolina (1), and Virginia (1). The data do not record facilities operating in 1907 in the remaining 21 states.

```{r bsmap, fig.cap = "Boarding schools in operation in 1907 by state"}
bs_ts<-bs %>% 
  left_join(st)
ggplot(bs_ts %>% 
         filter(year == 1907),
       aes(fill = boarding_schools,
           geometry = geometry)) + 
  geom_sf() + 
  scale_fill_viridis_c(trans = "sqrt") + 
  theme_void() + 
  labs(fill = "Facilities\noperating") + 
  theme(legend.position = "bottom")
```

## Model results: relationships between settler colonial projects and social policy infrastructure

We describe the results of our regression models in terms of expected values of each infrastructure outcome described above. We compute mean expected values and 95 percent credible intervals for each decennial year 1900 - 1930 from model posterior distributions for each outcome across three scenarios that closely correspond to the regional geography of the US during this period. We describe these scenarios in Table \@ref(tab:scen). All models include state-specific intercepts and slopes over time to account for other unmeasured features of states that impact both levels of employment and rates of change in employment.

The "High land redistribution" scenario is intended to describe states with a high volume of lands patented under the Morill and Homestead acts. These states typically had relatively low levels of Black tenant farmers per capita and white urban immigrants per capita (see Figure \@ref(fig:three-maps). For this scenario, we fix the sum of cumulative Homestead Act and Morill Act acres at the 90th percentile of the observed data (12.5 million acres), Black tenant farm residents per capita at the 5th percentile of the observed data (about 2 per 100,000 population), and white urban immigrants per capita at the 5th percentile of the observed data (about 2.6 per 1,000 population). This scenario is similar to the observed values of North Dakota, South Dakota, Montana, Colorado, Wyoming, and Kansas. 

The "High sharecropping scenario" is intended to describe states with a high volume of Black residents on tenant farms. These states, all in the US South, typically had very small urban populations and did not see a high volume of lands parceled under the Homestead and Morill acts. For this scenario, we fix Black tenant farm residents per capita at the 90th percentile of the observed data (143 per 1,000 population), white urban immigrant population per capita at the 5th percentile of the observed data (2.6 per 1,000 population), and the sum of Homestead and Morill acres patented at the 5th percentile of the observed data (0 acres). This scenario is most similar to Georgia and South Carolina during this period. The scenario is also similar to Louisiana, Alabama and Mississippi, though these states did have substantial acres of land patented under the Homestead and Morill acts prior to 1900. 

```{r scen, results = "asis", tab.cap= "Scenarios specified for model predictions"}
year_c = 0:3
STUSPS <- "XX"
tot_pop = 1e3

pred_dat<-data.frame(
  scen = c("High sharecropping", "High urban immigrants", "High land redistribution"),
  blk_sharecrop_pc = quantile(mdat$blk_sharecrop_pc, c(0.9, 0.05, 0.05)),
  urban_immig_pc = quantile(mdat$urban_immig_pc, c(0.05, 0.9, 0.05)),
  homestead_plus_morill = quantile(mdat$homestead_plus_morill, c(0.05, 0.05, 0.9))) %>% 
  expand_grid(year_c, tot_pop)

tab_dat<-pred_dat %>% 
  filter(year_c == 0) %>% 
  distinct() %>% 
  mutate(blk_sharecrop_pc = blk_sharecrop_pc * 1e3,
         urban_immig_pc = urban_immig_pc * 1e3,
         homestead_plus_morill = homestead_plus_morill / 1e6) %>% 
  select(-year_c, -tot_pop) %>% 
  rename("Scenario" = scen,
         "Black tenant\nfarmers per 1,000 pop." = blk_sharecrop_pc,
         "White urban\nimmigrants per 1,000 pop." = urban_immig_pc,
         "Homestead + Morill\nacres patented" = homestead_plus_morill)

kable(tab_dat, digits = 2)
```

The "High urban immigrants" scenario corresponds to state with large urban industrial centers that experienced a high volume of European immigration between 1900 and 1930. These states typically had very low numbers of Black tenant farmers during this period. States in the Northeast did not see lands patented to settlers under the Homestead and Morill acts. However, states on the West Coast and Midwest, some of which had large urban immigrant populations, were exposed to high-levels of homesteading and settler land redistribution during this period. For ease of comparison, we set the "High urban immigrant" scenario at the 90th percentile of observed white urban immigrants per capita (176 per 1,000 residents), and the 5th percentile of observed Black tenant farm residents per capita and cumulative Homestead and Morill acres patented. This scenario is similar to conditions in Connecticut, Illinois, Massachusetts, New Jersey, New York, and Rhode Island. 

In Figure \@ref(fig:emp-m-vis) we display our estimates of the spatial and temporal relationships between employment in key social policy professions and settler colonial project indicators. In Figure \@ref(fig:gq-m-vis) we show our estimates for the relationship between institutional group quarters populations and settler colonial project indicators. And in Figure \@ref(fig:bs-m-mvs) we show our estimates of the spatial and temporal relationships between operating Federal Indian boarding schools and these measures of settler colonial projects. 

In states where Native lands were being actively redistributed to settlers, there were higher levels of employment of teachers, and relatively lower levels of employment of police and social workers. These states had per capita rates of detention in correctional facilities, mental institutions, and institutions for the poor and disabled that were near the national average. They were dramatically more likely than other states to build and maintain Indian boarding schools. 

In states where there were large numbers of white immigrants residing in cities, teacher employment was around the national average, but police and social worker employment were dramatically higher than in other states. Rates of detention in correctional institutions were similar to all other states at the time, but the rate of detention in institutions for the mentally ill, poor, or disabled were much higher in these states than in other parts of the country. And while there were notable examples, such as the notorious Carlisle school in Pennsylvania, states with large urban white immigrant populations did not routinely build and maintain Indian boarding schools

In states with large numbers of Black tenant farmers between 1900 and 1930, there was much lower employment of teachers, police officers, and social workers than in other US states. Despite these low levels of employment in front-line social policy positions, however, these states had similar rates of incarceration in correctional institutions compared to other states. They had relatively lower levels of detention in institutions for the mentally ill, the poor, or the disabled. These states were also the least likely to build and maintain Indian boarding schools when compared with all other US states. 

### Employment

There was a dramatic increase in employment in education, policing, and social welfare between 1900 and 1930 nationally (see Figure \@ref(fig:emp-ts)), with substantial differences across regions of the US (see Figure \@ref(fig:emp-maps)). We construct a series of regression models to explore how the geography of settler projects related to both levels of employment and growth in employment across states. We display key results from these models as posterior expected values for focal scenarios in Figure \@ref(fig:emp-m-vis). We find that there was substantial growth in employment across all settler project contexts between 1900 and 1930, initial levels of employment and rates of change in employment varied dramatically across contexts.

```{r emp-m-vis, fig.cap = "Posterior expected employment rates per 1,000 population by employment type across settler project contexts, US States 1900 - 1930. Mean posterior expected value (line) and 95 percent credible interval (shaded)."}
preds_teach<-teachers_m1 %>% 
  add_epred_draws(re_formula = NA,
                  newdata = pred_dat) %>% 
  mutate(type = "Teachers")

preds_cops<-cops_m1 %>% 
  add_epred_draws(re_formula = NA,
                  newdata = pred_dat)%>% 
  mutate(type = "Police")

preds_socwork<-socwork_m1 %>% 
  add_epred_draws(re_formula = NA,
                  newdata = pred_dat)%>% 
  mutate(type = "Social workers")
  
p1<-ggplot(preds_teach,
       aes(x = year_c*10 + 1900, y = .epred)) + 
  stat_lineribbon(show.legend = F, 
                  .width = 0.95, 
                  fill = "gray50",
                  linewidth = 0.5) +
  theme_bw() + 
  theme(panel.spacing = unit(0.4, "cm")) +
  facet_wrap(~scen) + 
  labs(y = "",
       x = "",
       subtitle = "1. Teachers") 

p2<-ggplot(preds_cops,
       aes(x = year_c*10 + 1900, y = .epred)) + 
  stat_lineribbon(show.legend = F, 
                  .width = 0.95, 
                  fill = "gray50",
                  linewidth = 0.5) +
  theme_bw() + 
  theme(panel.spacing = unit(0.4, "cm")) +
  facet_wrap(~scen) + 
  labs(y = "Employment rate",
       x = "",
       subtitle = "2. Police") 

p3<-ggplot(preds_socwork,
       aes(x = year_c*10 + 1900, y = .epred)) + 
  stat_lineribbon(show.legend = F, 
                  .width = 0.95, 
                  fill = "gray50",
                  linewidth = 0.5) +
  theme_bw() + 
  theme(panel.spacing = unit(0.4, "cm")) +
  facet_wrap(~scen) + 
  labs(y = "",
       x = "Year",
       subtitle = "3. Social workers") 

grid.arrange(p1, p2, p3, nrow = 3)
```

Across categories of employment, we find clear relationships between settler colonial contexts and levels of employment in welfare and carceral institutions. We argue that the rationally and temporally varying interests of settler elites determine the kinds of colonial projects they prioritize, reflected in the kinds of human infrastructure that they develop to pursue policy goals. 

#### Teachers

```{r callouts1}
temp <- preds_teach %>% 
  group_by(year_c, scen) %>% 
  summarize(yhat_mn = mean(.epred), 
            yhat_lwr = quantile(.epred, 0.025), 
            yhat_upr = quantile(.epred, 0.975)) %>% 
  arrange(scen, year_c)
```

Panel 1 of Figure \@ref(fig:emp-m-vis) shows posterior expected values and 95 percent credible intervals for teacher employment across contexts of high settler land redistribution, high levels of sharecropping, and high levels of white urban immigrant population composition for each decennial Census year 1900 - 1930. Across all contexts, our models show substantial heterogeneity in baseline state employment levels ($\sigma_\gamma = 0.23 [0.19, 0.29]$) and in growth trajectories ($\sigma_\delta = 0.06 [0.04, 0.08]$).

We find that states with higher levels of settler land redistribution under the Homestead and Morill acts employed more dramatically more teachers per capita than did other states between 1900 and 1930. Model results suggest that states undergoing high volumes of settler land redistribution in 1900 had expected teacher employment rates near 7.2 [CI 6.3, 8.1] per 1,000 persons. By 1930, expected teacher employment would increase to 10.8 per 1,000 [10.2, 11.4]. This represented a nearly 50 percent increase in teacher employment over this period of 30 years. By 1930, we estimate that high land redistribution states employed teachers at a per capita rate about 40 percent higher than high sharecropping states and about 30 percent higher than high white urban immigrant states.

States with high levels of Black tenant farm residency had lower expected teacher employment than did other states. Our model suggests that a high sharecropping state could be expected to have about 3.9 [3.5, 4.4] teachers per 1,000 population in 1900, growing to about 7.5 [6.9, 8.0] teachers per 1,000 in 1930, about a 90 percent increase. We find evidence that growth in teacher employment in high sharecropping states was higher than in other contexts. Despite this higher-than-average growth, however, expected teacher employment in high sharecropping states in 1930 remained lower on average than in states that did not have widespread use of the sharecropping system. Our models suggest that by 1930, high sharecropping states had just caught up to teacher employment levels common in high land redistribution states in 1900.

In states with a high proportion of their population composed of white urban immigrants, our models suggest an expected teacher employment rate of 5.8 [5.2, 6.4] per 1,000 in 1900 and a rate of 8.1 [7.5, 8.7] in 1930, an increase of about 40 percent. Our models suggest that these states had slightly lower rates of growth in teacher employment than the national average rate of growth. 

#### Police

```{r callouts2}
temp <- preds_cops %>% 
  group_by(year_c, scen) %>% 
  summarize(yhat_mn = mean(.epred), 
            yhat_lwr = quantile(.epred, 0.025), 
            yhat_upr = quantile(.epred, 0.975)) %>% 
  arrange(scen, year_c)
```

Panel 2 of Figure \@ref(fig:emp-m-vis) shows expected values and credible intervals from our model of police employment across contexts. As with our teacher employment model, we find considerable variation in baseline state values ($\sigma_\gamma = 0.33 [0.26, 0.43]$) and in rates of growth ($\sigma_\delta = 0.05 [0.02, 0.07]$). 

State with large populations of white urban immigrants had much higher police employment levels than did other states during this period. In 1900, our model suggests an expected police employment rate of about 8.0 [6.6, 9.5] per 10,000 population in these states. This employment rate increased to about 15.5 [13.7, 17.5] per 10,000 by 1930, about a 94 percent growth in the expected employment rate over these three decades. We find no clear evidence that the proportional growth rates in police employment in these states were higher than the national rate of growth. Our results suggest that by 1930, these states would employ police at a rate nearly double the rate of police employment in high sharecropping states and high land redistribution states.

In states with high levels of settler land redistribution, our models suggest average police employment rates of about 3.4 per 10,000 in 1900 [2.7, 4.2], growing to about 8.7 [7.9, 9.7] per 10,000 in 1930. Police employment grew faster [2.5 percent, 12.2 percent] in these states than the national average. Our models suggest that a typical high settler land redistribution state had lower levels of police employment that other kinds of states in 1900, but by 1930, had higher levels of police employment than high sharecropping states, but still much lower employment levels than high urban immigrant states. 

In states with large populations of Black tenant farmers, we find comparatively low levels of police employment. In 1900, our model expects a high sharecropping state to employ 4.0 [3.4, 4.9] police per 10,000 population. In 1930, expected police employment increases to 7.1 [6.2, 8.0] per 10,000. We do not find evidence that the rate of change of police employment in these contexts is differs from other contexts during this period. This finding is somewhat surprising given the prominent role of law enforcement in cooperating with and upholding the regime of racial terror in force throughout the US South during this period, an issue we explore at greater length in the discussion.

#### Social workers

In Panel 3 of Figure \@ref(fig:emp-m-vis), we show model expectations for social work employment across contexts and over time. Social workers were very rarely documented in the Census in 1900. However, by the end of the period, there were substantial numbers of social workers across the US. As shown in Figure \@ref(fig:emp-ts), there was explosive growth in social worker employment during the Progressive Era, which varied dramatically across places ($\sigma_\gamma = 0.54, \sigma_\delta = 0.1$). We estimate a national expected growth rate of between 300 and 356 percent per decade. We do not detect meaningful differences in the growth rate of social workers across contexts, though there are dramatic differences in baseline levels, which have profound impacts on employment rate trajectories.

```{r callouts3}
temp <- preds_socwork %>% 
  group_by(year_c, scen) %>% 
  summarize(yhat_mn = mean(.epred) * 1e2, 
            yhat_lwr = quantile(.epred, 0.025)* 1e2, 
            yhat_upr = quantile(.epred, 0.975) * 1e2) %>% 
  arrange(scen, year_c)
```

Historians have long argued that states with large populations of white urban immigrants were the locus of the development of social work as a field (e.g. GORDON, WARD, POLSKY, ETC), and our findings strongly support this claim. We find that states with a large urban immigrant population had expected levels of social work employment (about 0.7 [0.4, 1.0] per 100,000 population) that were about 60 [27, 103] percent higher than other states in 1900. Our estimate of expected social work employment rises to 2.3 [1.7, 3.0] in 1910, 7.9 [6.0, 10.1] in 1920, and 27.2 [19.7, 36.6] in 1930, a forty-fold increase over 30 years. 

In line with the observed geography of employment shown in \ref(@fig:emp-maps), we find levels of social work employment were relatively low in states with high levels of settler land redistribution. Expected social work employment rise from about 0.2 [0.1, 0.3] per 100,000 in 1900 to about 6.1 [4.6, 7.9] per 100,000 in 1930. 

In states with large numbers of Black tenant farmers, social work employment rates began very near zero in 1900. Our expected social work employment rate in 1900 for a high sharecropping state is about 0.2 [0.1, 0.3] social workers per 100,000 population. By 1930, our analysis expects that a high sharecropping state would have about 5.4 [4.0, 7.3] social workers per 100,000 population, about a 30-fold increase. By 1930, we estimate that states with high numbers of urban white immigrants would have had between `r round(temp[12, 4] / temp[8, 5],1)` and `r round(temp[12, 5] / temp[8, 4],1)` times more social workers per capita than states with high levels of sharecropping. 

#### Indian boarding schools

As shown in Figure \@ref(fig:bsts), the count of boarding schools operating annually in the US fluctuated around 200 schools operating between 1900 and 1930. We estimate a logistic regression model to estimate the likelihood that a state had at least one boarding school in operation for each decade during the focal period. The posterior point estimates and credible intervals for focal parameters are shown in Figure \@ref(fig:bs-m).

```{r bs-m, fig.cap = "Relationships between settler colonial contexts and presence of Indian boarding schools. Logistic regression with operating boarding school binary outcome. Posterior mean and credible intervals (0.66, 0.95) for focal Beta parameters"}

pdat<-bs_m1_z %>%
  gather_draws(c(b_scaleblk_sharecrop_pc,
                 b_scaleurban_immig_pc,
                 b_scalehomestead_plus_morill)) %>% 
  mutate(.variable = case_when(
    .variable == "b_scaleblk_sharecrop_pc" ~ "Sharecropping",
    .variable == "b_scaleurban_immig_pc" ~ "Urban immigrants",
    .variable == "b_scalehomestead_plus_morill" ~ "Homestead + Morill acres"
  ))

ggplot(pdat,
       aes(x = .value,
           y = .variable)) +
  stat_pointinterval() +
  geom_vline(aes(xintercept = 0), lty = 2) +
  labs(x = "Estimate",
       y = "") +
  theme_bw()

```

```{r include = F}
# callouts
temp<-mdat %>% 
  filter(boarding_schools>0) %>% 
  select(STUSPS, year, homestead_plus_morill, boarding_schools) %>% arrange(homestead_plus_morill)

temp %>% filter(homestead_plus_morill==0)

# how many no morill+hs acre states
mdat %>% filter(homestead_plus_morill==0) %>% select(STUSPS) %>% distinct() %>% nrow()
### how many ANY morill + hs acre states
mdat %>% filter(homestead_plus_morill>0) %>% select(STUSPS) %>% distinct() %>% nrow()
# how many no bs with any morill + hs acres
mdat %>% filter(homestead_plus_morill>0 & boarding_schools ==0) %>% select(STUSPS) %>% distinct()

```

We find that states actively redistributing Native lands to settlers during the Progressive Era were dramatically more likely to operate boarding schools during than were other states. Among the 19 states that did not have lands parceled to settlers under the Homestead or Morill acts (though these states were unarguably beneficiaries of the acts), only New York, North Carolina, Pennsylvania, and Virginia housed Indian boarding schools during the Progressive Era. Among the 29 states within which Native lands were redistributed to settlers under the Morill or Homestead Acts, federal data suggest that only 6 did not have an operating Indian boarding school during this period. These patterns our clearly reflected in our regression results. States with high levels of sharecropping were less likely than states in other settler colonial contexts to operate a boarding school during the Progressive Era. White urban immigrant population does not have a meaningful statistical relationship to boarding school operation that we detect conditional on our data and model specifications. A large number of states with large urban populations operated no boarding schools, while one of the most notorious facilities, the Carlisle Industrial School, operated in Pennsylvania, among the leaders in white urban immigrant population during this period.

### Physical institutions

There are also clear relationships between settler colonial contexts and the magnitude of confinement in carceral and welfare institutions across states between 1900 and 1930. Figure \@ref(fig:gq-m-vis) displays posterior distributions of focal parameters for models of the populations housed in institutional group quarters between 1900 and 1930 by settler context scenario. 

By 1930, we find that expected rates of incarceration in prisons and jails were similar across each settler colonial context we specify. By contrast, we find lower expected levels of confinement in mental institutions in states with high levels of sharecropping compared to other contexts, and very high rates of mental institutionalization in places with a high volume of white urban immigrants. Similarly, we see lower rates of residence in institutions for the poor, elderly and disabled in high sharecropping states than in other contexts and much higher capacities in these institutions in high urban immigrant contexts.

NOTE FLAT / slight DOWN TRENDS IN NE STATES. INDICATES BUILDOUT HAPPENED EARLIER. BUILDOUT OCCURRING DURING PERIOD FOR OTHER STATES. 

<!-- CONSISTENT WITH HISTORICAL WORK ON PREVALENCE OF EARLY THERAPEUTIC / WELFARE STATES. CITE ALL THAT STUFF. These institutions reflected enlightenment ideals about the FOUCAULT AND STUFF. But they also operated within ideological contexts of rigid racist classification and eugenic policy priorities (METZL ETC). Where they existed, therapeutic and welfarist institutions routinely excluded Black people. These data further illustrate that the geographic distribution of these facilities followed similar racial logics. Welfare and mental institutions were relatively rare in the Jim Crow South, and very common in states with large numbers of white immigrants in cities. -->


<!-- Per capita levels of incarceration in correctional institutions were similar across contexts for all periods in our analyses, with relatively stable levels between 1900 and 1920, followed by an increase in incarceration rates in 1930. This homogeneity in carceral capacity is notable, given the differences in police employment described in Figure \@ref(fig:postpred-police). These findings illustrate that while states in the US South under Jim Crow and frontier states in the Midwest and West may have had relatively small police organizations compared to their peers in the urban Northeast and Midwest, the magnitude of their capacity for incarceration was similar. And while the absolute rate of incarceration was similar across states during this era, this should not diminish the importance of substantial differences in the conditions and contexts of confinement across places (CITE PRISON VARIATION LIT). Historical scholarship has clearly shown the salience of white supremacist and settler colonial logics for the development of early-modern policing and incarceration across US geographies (CITE CONVICT LEASE LIT, KLH ON California, NORTHEAST LIT, FRONTIER STUFF (MAYBE ROSS)). -->


```{r gq-vis, fig.cap = "Posterior expected employment rates per 1,000 population by employment type across settler project contexts, US States 1900 - 1930. Mean posterior expected value (line) and 95 percent credible interval (shaded)."}
preds_prison<-prison_m1 %>% 
  add_epred_draws(re_formula = NA,
                  newdata = pred_dat) %>% 
  mutate(type = "Prison")

preds_mental<-mental_m1 %>% 
  add_epred_draws(re_formula = NA,
                  newdata = pred_dat)%>% 
  mutate(type = "Mental inst.")

preds_poor<-poor_m1 %>% 
  add_epred_draws(re_formula = NA,
                  newdata = pred_dat)%>% 
  mutate(type = "Poverty inst.")

p1<-ggplot(preds_prison,
       aes(x = year_c*10 + 1900, y = .epred)) + 
  stat_lineribbon(show.legend = F, 
                  .width = 0.95, 
                  fill = "gray50",
                  linewidth = 0.5) +
  theme_bw() + 
  theme(panel.spacing = unit(0.4, "cm")) +
  facet_wrap(~scen) + 
  labs(y = "",
       x = "",
       subtitle = "1. Correctional institutions") 

p2<-ggplot(preds_mental,
       aes(x = year_c*10 + 1900, y = .epred)) + 
  stat_lineribbon(show.legend = F, 
                  .width = 0.95, 
                  fill = "gray50",
                  linewidth = 0.5) +
  theme_bw() + 
  theme(panel.spacing = unit(0.4, "cm")) +
  facet_wrap(~scen) + 
  labs(y = "Rate",
       x = "",
       subtitle = "2. Mental institutions") 

p3<-ggplot(preds_poor,
       aes(x = year_c*10 + 1900, y = .epred)) + 
  stat_lineribbon(show.legend = F, 
                  .width = 0.95, 
                  fill = "gray50",
                  linewidth = 0.5) +
  theme_bw() + 
  theme(panel.spacing = unit(0.4, "cm")) +
  facet_wrap(~scen) + 
  labs(y = "",
       x = "Year",
       subtitle = "3. Institutions for poor, disabled, and elderly") 

grid.arrange(p1, p2, p3)
```

#### Prisons and jails

There was considerable variation in populations detained in correctional institutions across places and over time during the Progressive Era. We estimate large differences in average incarceration rates across places ($\sigma_\gamma = 0.31 [0.09, 0.54]$), but have a high level of uncertainty about any meaningful differences in state-level growth trajectories ($\sigma_\delta = 0.05 [0, 0.14]$). As shown in Figure \@ref(fig:gq-vis), incarceration rates trended upward in high sharecropping and high land redistribution contexts during this period. In contexts of large white urban immigrant populations, incarceration rates remained at relatively high levels throughout the period, despite a small decline in expected incarceration over these four decades. 

In state with large urban immigrant populations, our models suggest that incarceration rates could be expected to be in the range of about 1-2 per 1,000 population between 1900 and 1930. We estimate an expected incarceration rate of 1.7 [1.2, 2.3] for this scenario in 1900, much higher than expected incarceration rates in other contexts. This rate remains relatively stable over the full period with a small expected decline to a rate of 1.2 [0.9, 1.7] by 1930. 

In states with high levels of Native land redistribution, we see a substantial rise in expected incarceration. In 1900, we expect a state with high levels of Morill and Homestead acres patented to have an incarceration rate of about 0.6 [0.3, 1.0] per 1,000 population. This expected value rises rapidly to 0.85 [0.6, 1.2] in 1910, 1.2 [0.9, 1.6] in 1920 and 1.8 [1.3, 2.4] in 1930. This growth represents about 300 percent growth in expected incarceration rates. By 1930, these states had higher expected incarceration rates than other scenarios, though there is a substantial amount of uncertainty in the rank order of scenarios for 1930. 

States with a large population of Black tenant farmers also saw substantial growth in incarceration rates during this period. These states also started with significant prison and jail capacities. We estimate an expected incarceration rate of 0.8 [0.6, 1.1] per 1,000 population in high sharecropping states in 1900. By 1930, expected incarceration rates rose by about 90 percent by 1930 to a rate of 1.5 [1.1, 2.2]. Despite relatively thin infrastructure on other measures, states in the Jim Crow South had incarceration rates that were similar to the rest of the country. Carceral investments appear to be an exception to the trend of under-development in other areas of social infrastructure in high sharecropping states. 

```{r callouts4}
temp_prison <- preds_prison %>% 
  group_by(year_c, scen) %>% 
  summarize(yhat_mn = mean(.epred) , 
            yhat_lwr = quantile(.epred, 0.025), 
            yhat_upr = quantile(.epred, 0.975)) %>% 
  arrange(scen, year_c)

temp_mental <- preds_mental %>% 
  group_by(year_c, scen) %>% 
  summarize(yhat_mn = mean(.epred) , 
            yhat_lwr = quantile(.epred, 0.025), 
            yhat_upr = quantile(.epred, 0.975)) %>% 
  arrange(scen, year_c)

temp_poor <- preds_poor %>% 
  group_by(year_c, scen) %>% 
  summarize(yhat_mn = mean(.epred) , 
            yhat_lwr = quantile(.epred, 0.025), 
            yhat_upr = quantile(.epred, 0.975)) %>% 
  arrange(scen, year_c)
```

#### Mental institutions

There are clear and striking regional patterns in the distribution of the US population housed in mental institutions across states during the Progressive Era. As shown in Figure \@ref(fig:gq-vis) Panel 2, states with large white urban immigrant populations had much higher levels of confinement in mental institutions than did other states. However, these levels appear to have remained flat during this 40-year period. Mental institutionalization grew rapidly in states with high levels of settler land redistribution. Capacity in mental institutions appears to have grown, albeit at a slow rate, in high sharecropping states. We find substantial variation in individual state intercepts ($\sigma_\gamma = 0.32 [0.07, 0.56]$) but thin evidence for heterogeneity in state-specific growth trajectories ($\sigma_\delta = 0.05[0.00,0.14]$).

We find that states actively redistributing Native lands to settlers have an expected mental institutionalization rate of about 1.5 [0.9, 2.4] per 1,000 population in 1900. This rate rose dramatically during the Progressive era. Mean expected values rose by about 290 percent between 1900 and 1930, ending the period at a rate of about 4.5 [3.6, 5.6] per 1,000. This rapid growth suggests substantial investments in expanding capacity for mental institutionalization in states undergoing rapid settler land redistribution in the early 20th century. 

States in the sharecropping South tended to have lower levels of mental institutionalization than other regions. In 1900, rates of institutionalization were relatively low, at 1.89 [1.4, 2.5] per 1,000 population, similar to levels observed in high land redistribution states. However, rates of institutionalization grew much less quickly in high sharecropping states than in high land redistribution states. By 1930, these states had an expected rate of mental institutionalization of 3.2 [2.5, 4.0], somewhat lower than high sharecropping states and substantially lower than high urban immigrant population states. Like other paternalistic welfare and public health institutions, states in the US South appear to have invested far fewer resources in infrastructure when compared with peer states (UNDERDEVELOPMENT CITES?).

#### Institutions for the poor, disabled and elderly

As with mental institutions, we find large expected increases in populations housed in institutions for the poor, elderly, and disabled in states with high levels of settler land redistribution between 1900 and 1930, relatively small increases in size in high sharecropping states, and small decreases in already-high levels in states with large numbers of white urban immigrants. We find substantial heterogeneity in both intercepts ($\sigma_\gamma = 0.6[0.39, 0.83]$ and in their growth trajectories $[\sigma_\delta = 0.09[0.01. 0.18]$).

High sharecropping states in the South tended to have very low levels of populations in these kinds of institutions across the period. In 1900, we estimate an expected rate of 0.96 [0.6, 1.4] residents per 1,000 population housed in institutions for the poor, disabled and elderly. This rate grew slowly over the 40 year period, to and avergage of 1.1 in 1910, 1.3 in 1920, and 1.5 [1.1, 2.1] in 1930. Again, these results indicate generally lower levels of investment in social institutions in the Jim Crow South than in other settler contexts.

We find that states with large populations of urban immigrants had relatively large numbers of people housed in poverty institutions like nursing homes, poorhouses, and orphanages. In 1900, these states have an expected rate of 4.3 [2.7, 6.3] per 1,000 housed in poverty institutions. By 1930, this rate had fallen slightly to 3.7 [2.6, 5.1] per 1,000. Despite this decline, these states had higher levels of housing in poverty institutions in 1930 than did states in our other two focal scenarios.  

As with other infrastructure outcomes, we see a rapid growth in the population of people housed in poverty institutions in states with high levels of settler land redistribution. We expect a rate of about 0.79 [0.4, 1.4] per 1,000 population in these facilities in 1900, rising to about 2.6 [1.9, 3.4] per 1,000 in 1930. These states had a rate of growth that was significantly faster than states enmeshed in other settler colonial projects during the Progressive Era. 

